\clearemptydoublepage

\phantomsection
\addcontentsline{toc}{chapter}{Outline of the Thesis}

\begin{center}
	\huge{Outline of the Thesis}
\end{center}




%--------------------------------------------------------------------
\section*{Part I: Introduction and Theory}

\noindent {\scshape Chapter 1: Introduction}  \vspace{1mm}

\noindent  This chapter presents a general introduction and motivation to the topic of parallel and general purpose GPU computing.\\

\noindent {\scshape Chapter 2: Theory}  \vspace{1mm}

\noindent  This chapter introduces the relevant technologies that are used in the implementation, namely OpenCL and FunkyIMP.   \\

%--------------------------------------------------------------------
\section*{Part II: Methodology}

\noindent {\scshape Chapter 3: Extensions to the funkyIMP Compiler}  \vspace{1mm}

\noindent  This chapter presents the extensions that were made to the funkyIMP compiler in order to accommodate the automated compilation of code on the GPU.   \\

\noindent {\scshape Chapter 4: The Benchmark Suite}  \vspace{1mm}

\noindent  This chapter presents the benchmark suite that was written to create a model of the runtime of the individual operations in OpenCL.   \\

%--------------------------------------------------------------------
\section*{Part III: Runtime Model and Results}

\noindent {\scshape Chapter 5: A Runtime Prediction Model}  \vspace{1mm}

\noindent  This chapter introduces presents the actual runtime model for OpenCL operations.   \\

\noindent {\scshape Chapter 6: Results and Evaluation}  \vspace{1mm}

\noindent  In this chapter, the results achieved in the course of this thesis are shown. The quality of the predictions made utilizing the runtime prediction model is evaluated on a quantitative scale.\\

%--------------------------------------------------------------------
\section*{Part IV: Further Work and Conclusion}

\noindent {\scshape Chapter 7: Future Work}  \vspace{1mm}

\noindent  There are topics that did not fit in the scope of this thesis. This chapter discusses several of these.   \\

\noindent {\scshape Chapter 8: Conclusion}  \vspace{1mm}

\noindent  In this chapter the results of this thesis are evaluated on a qualitative scale.  \\